---
title: "Public Pension Plan Prototypes"
author: "Don Boyd"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
---

```{r runall, eval=FALSE, echo=FALSE}
# When we want a final report, run the following code selectively "by hand" (interactively) -- NEVER using Knit with eval=TRUE

rmdfn <- "./_Main.rmd" # this file
outfn <- paste0("./Results/", "PPPPrototypes_", format(Sys.time(), "%Y-%m-%d"), ".html")
rmarkdown::render(rmdfn, output_format="html_document", output_file=outfn)

# Note that it is safest to fully exit RStudio and restart it before running the whole thing. Otherwise knitr can get confused
# and include repetitive information in the output html file.

```


```{r mainSet options, echo=FALSE, cache=FALSE}
options(width=120)
knitr::opts_chunk$set(fig.width=14, fig.height=10, echo=FALSE)
# Note: when saving maps (ggsave), width=16, height=9 seems to get rid of white space

```


```{r libs, message=FALSE}
library(dplyr)
options(dplyr.print_min = 60) # default is 10
options(dplyr.print_max = 60) # default is 20

library(knitr)
library(lubridate) 
library(ggplot2)
library(magrittr)
library(readr)
library(readxl)
library(stringr)
library(tidyr)

# load my packages last
#devtools::install_github("donboyd5/btools")
#devtools::install_github("donboyd5/pdata")
library(btools)
library(pdata) # pension data

# CRR locations for CAFRs and AVs
# http://publicplansdata.org/reports/  for all
# http://publicplansdata.org/resources/download-avs-cafrs/  menu system


```


```{r defines}
resultsdir <- "./Results/"

```


```{r makevars}
# create a df from the ppd, with constructed vars needed for prototype analysis

# start by setting up tools to find needed variables, given how many are in the ppd
ppdvars <- ppdvars # now we can look at it in the viewer
# glimpse(ppd)
# names(ppd)
findvars <- function(text) {
  # note that there may be some discrepancies between variables found in the "variables" file
  # and those actually in the data
  df <- ppdvars %>% filter(grepl(text, `Variable Label`, ignore.case=TRUE) | 
                             grepl(text, `Variable Name`, ignore.case=TRUE)) %>% 
    select(`Variable Label`, `Variable Name`)
  return(df)
}
# grep("salar", names(ppd), ignore.case=TRUE, value=TRUE) # search for variable names with desired text - use actual data file
# temp <- findvars("salar") # now use viewer to look at temp

# summary(filter(select(ppd, fy, ActFundedRatio_GASB), fy==2013)) # look for NAs by year

# variable choices when there are multiple options (results from analysis that isn't always shown below)
# assets: many NA for MktAssets_ActRpt; MktAssets_net looks better
# age variables: there are many: ActiveAge_avg, BeneficiaryAge_avg, ServiceRetireeAge_avg, ServiceRetAge_avg
#   but ALL have many missing values
#   ActiveAge_avg has fewest missing values - 110 not NA's in the median of 2009-2013; others are much worse
# so, do not use age in the analysis, but do look at it by cluster after plans are clustered

# another way to get at this may be through pvfb
# PVFB_active, PVFB_retiree
# explore:
# ppd %>% select(fy, ppd_id, contains("PVFB")) %>%
#   gather(variable, value, -fy, -ppd_id) %>%
#   group_by(variable, fy) %>% 
#   mutate(value=cton(value)) %>% 
#   do(qtiledf(.$value)) %>%
#   data.frame
# almost all values are missing; a few exceptions:
#  - we have 119 non-missing PVFB-active in 2012; not really as many as I would like
#  - PVFB-retiree is pretty good in most years, including 140 nonmissing in 2012
# so best we could do is retire/active ratio for maybe 119 plans
# ppd %>% mutate(rapvfb=PVFB_retiree/PVFB_active) %>%
#   group_by(fy) %>%
#   do(qtiledf(.$rapvfb))
# some big differences but probably not usable in clustering

# Classifiers:
# PlanType 1-PERS or SERS, 2- TRS, 3- Police/Fire/Safety
# AdministeringGovt  0-State, 1-County, 2-City, 5- School
ptlevs <- c(1, 2, 3); ptlabs <- c("General emp", "Teachers", "Safety")
adlevs <- c(0, 1, 2, 5); adlabs <- c("State", "County", "City", "School")


# get the desired data and calc vars
df <- ppd %>% select(ppd_id, PlanName, fy, ActFundedRatio_GASB, InvestmentReturnAssumption_GASB, PercentReqContPaid,
                     NormCostRate_tot, actives_tot, MktAssets_net,
                     contrib_tot, benefits_tot, expense_net, actives_tot, beneficiaries_ServiceRetirees,
                     ActiveAge_avg, ActiveSalary_avg) %>%
  group_by(ppd_id, PlanName) %>%
  mutate(activepch=actives_tot / actives_tot[match(fy, fy-1)] * 100 - 100, 
         xcfpct=(contrib_tot - benefits_tot) / MktAssets_net * 100,
         xcfpct2=(contrib_tot + expense_net) / MktAssets_net * 100, # see missing value analysis below for reasoning here
         abratio=actives_tot / beneficiaries_ServiceRetirees) %>%
  filter(fy>=2009, fy<=2013) %>% # there is some 2014 data in the file, but it is not ready yet
  summarise_each(funs(median(., na.rm=TRUE)), ActFundedRatio_GASB, InvestmentReturnAssumption_GASB,
                 PercentReqContPaid, NormCostRate_tot, activepch, xcfpct, xcfpct2, abratio,
                 ActiveAge_avg, ActiveSalary_avg)

```

# Investigate the data

## Simple distributions and correlations for plan-specific medians computed over the period 2009-2013
```{r simple}
# get distributions and # NAs
df %>% select(-PlanName) %>%
  gather(variable, value, -ppd_id) %>% 
  group_by(variable) %>%
  do(qtiledf(.$value)) %>%
  kable(digits=4, caption="Quantiles")
# good, xcfpct2 looks good
# active salary also has enough obs

# look at correlations among vars
cor(select(ungroup(df), -ppd_id, -PlanName), use="pairwise.complete.obs") %>% 
  kable(caption="Correlations, pairwise.complete.obs", digits=4)

```




## Missing values check
```{r mv, eval=FALSE}
# don't normally include this in the output
# which plans have missing data in vars we care about?
df %>% filter(is.na(NormCostRate_tot) | is.na(xcfpct)) %>% 
  select(ppd_id, PlanName, NormCostRate_tot, xcfpct) %>% # keep only selected vars so we have room to see results
  kable(digits=4) 
# normcost loses NYCFire, NYCPolice
# xcfpct loses 25, including CalSTRS, Ohio, Chicago Municipal, and Boston - that's too much to lose


# could we use expense_net instead of benefits_tot?
ppd %>% select(ppd_id, PlanName, fy, benefits_tot, expense_net) %>%
  group_by(fy) %>%
  summarise_each(funs(n(), nna=sum(is.na(.)), median(., na.rm=TRUE)), benefits_tot, expense_net)
# far fewer missing for expense than benefits; sign is negative

# how correlated are they?
cor(ppd[, c("benefits_tot", "expense_net")], use="complete.obs") # only .085
f <- function(v1, v2) {
  d <- cor(cbind(v1, v2), use="complete.obs")
  return(d[1, 2])
}
ppd %>% select(ppd_id, PlanName, fy, benefits_tot, expense_net) %>%
  group_by(fy) %>%
  summarise(cor=f(benefits_tot, expense_net)) # hmmm, correlation is .99 in most years except 2008
# look at 2008 - some digits appear way off eg Missouri local
tmp <- ppd %>% select(ppd_id, PlanName, fy, benefits_tot, expense_net) %>%
  filter(fy==2008) %>%
  mutate(ratio=benefits_tot / abs(expense_net)) %>%
  arrange(desc(ratio))
# MO Local appears to be the main problem in 2008; I've notified CRR and they say they have fixed it

ppd %>% filter(ppd_id==61) %>% 
  select(ppd_id, PlanName, fy, benefits_tot, expense_net) %>%
  mutate(ratio=benefits_tot / abs(expense_net)) %>%
  arrange(fy)

```


# Cluster analysis
```{r cluster}

df2 <- df %>% select(ppd_id, PlanName, ActFundedRatio_GASB, InvestmentReturnAssumption_GASB, PercentReqContPaid,
                     NormCostRate_tot, activepch, xcfpct2, abratio, ActiveSalary_avg) 

df.prescale <- na.omit(df2) # listwise deletion of missing (NYCPolice, NYCFire)
df.scaled <- df.prescale
df.scaled[, -c(1, 2)] <- scale(df.scaled[, -c(1, 2)]) # scale to mean=0, sd=1 before doing distance calculations
# verify mean sd
# df.scaled %>% gather(variable, value, -ppd_id, -PlanName) %>%
#   group_by(variable) %>%
#   summarise(mean = mean(value), sd = sd(value))
d <- dist(df.scaled[, -c(1:2)])   # calc Euclidean distance matrix 
hc <- hclust(d) # hierarchical cluster
plot(hc, labels=as.character(df.scaled$ppd_id)) # plot the dendrogram
print(hc)

# which plans in each cluster are closest to the center of their cluster? These plans might be used as a model for that kind of plan


```


## Summary of clusters
```{r}
# create a specific number of clusters and look at their characteristics
group <- cutree(hc, k=8) # assign each plan to a group
# put group id on the cluster variables df, and bring in other data
df.clust <- cbind(group, df.prescale) %>% 
  left_join(ppd %>% filter(fy==2013) %>% select(ppd_id, MktAssets_net, PlanType, AdministeringGovt)) %>%
  mutate(planf=factor(PlanType, levels=ptlevs, labels=ptlabs),
         adminf=factor(AdministeringGovt, levels=adlevs, labels=adlabs)) %>%
  select(-PlanType, -AdministeringGovt) %>%
  select(ppd_id, PlanName, planf, adminf, everything()) %>%
  arrange(group, desc(MktAssets_net))

count(df.clust, group) %>% arrange(desc(n))
clusterinfo <- df.clust %>% group_by(group) %>% 
  summarise_each(funs(median(., na.rm=TRUE)), -c(ppd_id, PlanName, planf, adminf))
# ActFundedRatio_GASB, PercentReqContPaid, NormCostRate_tot, activepch, xcfpct2, abratio, ActiveSalary_avg
cnt <- count(df.clust, group)
left_join(cnt, clusterinfo) %>% arrange(desc(n)) %>% 
  mutate(MktAssets_net=MktAssets_net/1e6) %>%
  kable(caption="Medians for each cluster", digits=3)

df.clust %>% group_by(group, planf) %>%
  summarise(n=n()) %>%
  mutate(planf=ifelse(is.na(planf), "Missing", as.character(planf))) %>%
  spread(planf, n, fill=0) %>%
  kable(caption="# of plans in each cluster, by plan type")


df.clust %>% group_by(group, adminf) %>%
  summarise(n=n()) %>%
  mutate(adminf=ifelse(is.na(adminf), "Missing", as.character(adminf))) %>%
  spread(adminf, n, fill=0) %>%
  kable(caption="# of plans in each cluster, by administration")


```


Comments:  

* 1 and 2 seem pretty similar - fairly average, underfunded plans
* 3 expensive relatively low outflow plans
* 4 poorly funded high outflow mature plans with bad-behaving sponsors
* 5 well funded low outflow high actives plans
* 6 well funded low discount rate plans


## Plans in each cluster, sorted by asset size
```{r}
f <- function(groupnum) {
  df.clust %>% filter(group==groupnum) %>%
    kable(digits=4)
}
# groupnum <- 1
f(1)
f(2)
f(3)
f(4)
f(5)
f(6)
f(8)
f(7)

```


## Plans in each cluster, ranked by distance from cluster center
```{r}
# stack the cluster summaries (clusterinfo) and the clustered plans df.clust
cl2 <- clusterinfo %>% mutate(PlanName=paste0("Group ", group, " medians"), ppd_id=-group)

df.all <- bind_rows(cl2, df.clust)

getdist <- function(df, clustvars) {
  df <- as.data.frame(df) # need old fashioned data frame to put rownames on it
  rownames(df) <- df$ppd_id # I want rownames because it guarantees I can't mix up what the distances measure
  scaled <- scale(df[, clustvars])
  d <- as.matrix(dist(scaled), labels=TRUE)
  distdf <- data_frame(ppd_id=as.numeric(names(d[, 1])), dist=d[, 1])
  return(distdf)
}

cvars <- c("ActFundedRatio_GASB", "InvestmentReturnAssumption_GASB", "PercentReqContPaid", "NormCostRate_tot", "activepch",
           "xcfpct2", "abratio", "ActiveSalary_avg")
df.dist <- df.all %>% group_by(group) %>%
  arrange(ppd_id) %>%
  do(getdist(., cvars))

# merge the distances back against the raw (unscaled) data and examine results
df.all2 <- left_join(df.all, df.dist) %>% arrange(group, dist)

f <- function(groupnum) {
  df.all2 %>% filter(group==groupnum) %>%
    mutate(MktAssets_net=MktAssets_net/1e6) %>%
    select(-group) %>%
    select(ppd_id, PlanName, dist, everything()) %>%
    kable(digits=4, caption=paste0("Group: ", groupnum, ".  Assets $ billions"))
}
f(1)
f(2)
f(3)
f(4)
f(5)
f(6)
f(8)
f(7)


```




